\mysection{{\tt str}: regular expressions and high-level string processing }

\label{s:str}
\index{str (module)@\verb`str` (module)}%

\subsection*{Regular expressions }\begin{verbatim}
type regexp
\end{verbatim}
\begin{comment}
 The type of compiled regular expressions. 
\end{comment}
\begin{verbatim}
value regexp: string -> regexp
\end{verbatim}
\index{regexp@\verb`regexp`}%
\begin{comment}
 Compile a regular expression. The syntax for regular expressions
           is the same as in Gnu Emacs. The special characters are
           \verb$^.*+?[]\. The following constructs are recognized:
\\[\smallskipamount]          \verb.      matches any character except newline
\\[\smallskipamount]          \verb*      (postfix) matches the previous expression zero, one or
                    several times
\\[\smallskipamount]          \verb-      (postfix) matches the previous expression one or
                    several times
\\[\smallskipamount]          \verb?      (postfix) matches the previous expression once or
                    not at all
\\[\smallskipamount]          \verb[..]   character set; ranges are denoted with \verb-, as in \verba-z;
                    an initial \verb^, as in \verb^0-9, complements the set
\\[\smallskipamount]          \verb^      matches at beginning of line
\\[\smallskipamount]          \verb$      matches at end of line
\\[\smallskipamount]          \verb\|     (infix) alternative between two expressions
\\[\smallskipamount]          \verb\(..\) grouping and naming of the enclosed expression
\\[\smallskipamount]          \verb\1     the text matched by the first \verb\(...\) expression
                    (\verb\2 for the second expression, etc)
\\[\smallskipamount]          \verb\b     matches word boundaries
\\[\smallskipamount]          \verb\^, \$ quotes special characters. 
\end{comment}
\begin{verbatim}
value regexp_case_fold: string -> regexp
\end{verbatim}
\index{regexp_case_fold@\verb`regexp_case_fold`}%
\begin{comment}
 Same as \verbregexp, but the compiled expression will match text
           in a case-insensitive way: uppercase and lowercase letters will
           be considered equivalent. 
\end{comment}
\subsection*{String matching and searching }\begin{verbatim}
value string_match: regexp -> string -> int -> bool
\end{verbatim}
\index{string_match@\verb`string_match`}%
\begin{comment}
 \verbstring_match r s start tests whether the characters in \verbs
           starting at position \verbstart match the regular expression \verbr. 
\end{comment}
\begin{verbatim}
value search_forward: regexp -> string -> int -> int
\end{verbatim}
\index{search_forward@\verb`search_forward`}%
\begin{comment}
 \verbsearch_forward r s start searchs the string \verbs for a substring
           matching the regular expression \verbr. The search starts at position
           \verbstart and proceeds towards the end of the string.
           Return the position of the first character of the matched
           substring, or raise \verbNot_found if no substring matches. 
\end{comment}
\begin{verbatim}
value search_backward: regexp -> string -> int -> int
\end{verbatim}
\index{search_backward@\verb`search_backward`}%
\begin{comment}
 Same as \verbsearch_forward, but the search proceeds towards the
           beginning of the string. 
\end{comment}
\begin{verbatim}
value matched_string: string -> string
\end{verbatim}
\index{matched_string@\verb`matched_string`}%
\begin{comment}
 \verbmatched_string s returns the substring of \verbs that was matched
           by the latest \verbstring_match, \verbsearch_forward or \verbsearch_backward.
           The user must make sure that the parameter \verbs is the same string
           that was passed to the matching or searching function. 
\end{comment}
\begin{verbatim}
value match_beginning: unit -> int
value match_end: unit -> int
\end{verbatim}
\index{match_beginning@\verb`match_beginning`}%
\index{match_end@\verb`match_end`}%
\begin{comment}
 \verbmatch_beginning() returns the position of the first character
           of the substring that was matched by \verbstring_match,
           \verbsearch_forward or \verbsearch_backward. \verbmatch_end() returns
           the position of the character following the last character of
           the matched substring. 
\end{comment}
\begin{verbatim}
value matched_group: int -> string -> string
\end{verbatim}
\index{matched_group@\verb`matched_group`}%
\begin{comment}
 \verbmatched_group n s returns the substring of \verbs that was matched
           by the \verbnth group \verb\(...\) of the regular expression during
           the latest \verbstring_match, \verbsearch_forward or \verbsearch_backward.
           The user must make sure that the parameter \verbs is the same string
           that was passed to the matching or searching function. 
\end{comment}
\begin{verbatim}
value group_beginning: int -> int
value group_end: int -> int
\end{verbatim}
\index{group_beginning@\verb`group_beginning`}%
\index{group_end@\verb`group_end`}%
\begin{comment}
 \verbgroup_beginning n returns the position of the first character
           of the substring that was matched by the \verbnth group of
           the regular expression. \verbgroup_end n returns
           the position of the character following the last character of
           the matched substring. 
\end{comment}
\subsection*{Replacement }\begin{verbatim}
value global_replace: regexp -> string -> string -> string
\end{verbatim}
\index{global_replace@\verb`global_replace`}%
\begin{comment}
 \verbglobal_replace regexp s repl returns a string identical to \verbs,
           except that all substrings of \verbs that match \verbregexp have been
           replaced by \verbrepl. The replacement text \verbrepl can contain
           \verb\1, \verb\2, etc; these sequences will be replaced by the text
           matched by the corresponding group in the regular expression.
           \verb\0 stands for the text matched by the whole regular expression. 
\end{comment}
\begin{verbatim}
value replace_first: regexp -> string -> string -> string
\end{verbatim}
\index{replace_first@\verb`replace_first`}%
\begin{comment}
 Same as \verbglobal_replace, except that only the first substring
           matching the regular expression is replaced. 
\end{comment}
\begin{verbatim}
value global_substitute: regexp -> (string -> string) -> string -> string
\end{verbatim}
\index{global_substitute@\verb`global_substitute`}%
\begin{comment}
 \verbglobal_substitute regexp subst s returns a string identical
           to \verbs, except that all substrings of \verbs that match \verbregexp
           have been replaced by the result of function \verbsubst. The
           function \verbsubst is called once for each matching substring,
           and receives \verbs (the whole text) as argument. 
\end{comment}
\begin{verbatim}
value substitute_first: regexp -> (string -> string) -> string -> string
\end{verbatim}
\index{substitute_first@\verb`substitute_first`}%
\begin{comment}
 Same as \verbglobal_substitute, except that only the first substring
           matching the regular expression is replaced. 
\end{comment}
\subsection*{Splitting }\begin{verbatim}
value split: regexp -> string -> string list
\end{verbatim}
\index{split@\verb`split`}%
\begin{comment}
 \verbsplit r s splits \verbs into substrings, taking as delimiters
           the substrings that match \verbr, and returns the list of substrings.
           For instance, \verbsplit (regexp "[ \t]+") s splits \verbs into
           blank-separated words. 
\end{comment}
\begin{verbatim}
value bounded_split: regexp -> string -> int -> string list
\end{verbatim}
\index{bounded_split@\verb`bounded_split`}%
\begin{comment}
 Same as \verbsplit, but splits into at most \verbn substrings,
           where \verbn is the extra integer parameter. 
\end{comment}
\subsection*{Joining }\begin{verbatim}
value concat: string list -> string
\end{verbatim}
\index{concat@\verb`concat`}%
\begin{comment}
 Catenate a list of string. 
\end{comment}
\begin{verbatim}
value join: string -> string list -> string
\end{verbatim}
\index{join@\verb`join`}%
\begin{comment}
 Catenate a list of string. The first argument is a separator, which
           is inserted between the strings. 
\end{comment}
\subsection*{Extracting substrings }\begin{verbatim}
value string_before: string -> int -> string
\end{verbatim}
\index{string_before@\verb`string_before`}%
\begin{comment}
 \verbstring_before s n returns the substring of all characters of \verbs
           that precede position \verbn (excluding the character at 
           position \verbn). 
\end{comment}
\begin{verbatim}
value string_after: string -> int -> string
\end{verbatim}
\index{string_after@\verb`string_after`}%
\begin{comment}
 \verbstring_after s n returns the substring of all characters of \verbs
           that follow position \verbn (including the character at 
           position \verbn). 
\end{comment}
\begin{verbatim}
value first_chars: string -> int -> string
\end{verbatim}
\index{first_chars@\verb`first_chars`}%
\begin{comment}
 \verbfirst_chars s n returns the first \verbn characters of \verbs.
           This is the same function as \verbstring_before. 
\end{comment}
\begin{verbatim}
value last_chars: string -> int -> string
\end{verbatim}
\index{last_chars@\verb`last_chars`}%
\begin{comment}
 \verblast_chars s n returns the last \verbn characters of \verbs. 
\end{comment}
\subsection*{Formatting }\begin{verbatim}
value format: string -> 'a
\end{verbatim}
\index{format@\verb`format`}%
\begin{comment}
 Same as \verbfprintf and \verbprintf from the \verbprintf module,
           except that the result of the formatting is returned as a
           string instead of being written on a channel. The structure
           of the format string is described in \verbprintf. 
\end{comment}
