% Copyright     Digital Equipment Corporation & INRIA     1988, 1989
% Last modified_on Wed Feb 27 19:49:15 GMT+1:00 1991 by herve

%\documentstyle[fullpageusa,makeidx]{article}    % Specifies the document style.
\documentstyle[makeidx]{article}

\title{{\Huge BigNum:  A Portable and \\ Efficient Package \\ for
Arbitrary-Precision \\ Arithmetic}}
\author{ B. Serpette  \thanks{Institut National de Recherche
          en Informatique et Automatique, 78150, Rocquencourt, France.} \\
        J. Vuillemin \thanks{Digital Equipment Corp., Paris Research Laboratory,
        85, Av. Victor Hugo.        92563 Rueil-Malmaison Cedex, France.}  \\
        J.C. Herv\'e  $^{\dagger}$  \\
       }  

\newtheorem{func}{Procedure}
\newcommand{\mod}{\bmod}
\newcommand{\lelisp}{LeLisp }
\newcommand{\ob}{\, _{b}[}
\newcommand{\N}{\mbox{{\bf N}}}
\newcommand{\n}{\mbox{{\verb+N+}}}
\newcommand{\m}{\mbox{{\verb+M+}}}
\newcommand{\p}{\mbox{{\verb+P+}}}
\newcommand{\q}{\mbox{{\verb+Q+}}}
\newcommand{\r}{\mbox{{\verb+R+}}}
\newcommand{\dd}{\mbox{{\verb+D+}}}
\newcommand{\bb}{\mbox{{\bf b}}}
\newcommand{\Z}{\mbox{{\bf Z}}}
\newcommand{\Bd}{\mbox{\bf B}_{2}}
\newcommand{\Bb}{\mbox{\bf B}_{b}}
\newcommand{\Nd}{\mbox{\bf N}_{2}}
\newcommand{\V}{{\cal V}}
\newcommand{\ips}[1]{\dotfill\ {\it #1}}

\makeindex
\begin{document}
              
\maketitle                 % Produces the title.
\begin{abstract}
We describe a C package for arbitrary-precision integer
arithmetic that is portable, yet efficient. Making the package run fast
on a given computer involves re-writing a small
kernel of our package in native assembly language. 
We provide such assembly code for VAX, 68020 and NS instruction sets,
with good benchmarks. This package serves as a foundation for
two arithmetic packages written in higher level
languages \lelisp \footnote{\lelisp is a registered
trade-mark of INRIA.} and Modula2+. The package is publicly available for
non-commercial uses.
This report will also appear in French as an INRIA report.
\end{abstract}

\clearpage

\newpage

\input{bnbody.tex}

\clearpage

{\Large \bf Acknowledgements}\\

We are thankfull to the following people, for their contributions
to this package:
Patrice Bertin, Hans Boehm, Jerome Chailloux, Michel Gangnet, 
Fran\c{c}ois Morain, David Salesin and Mark Shand.

\vspace{2cm}

\tableofcontents

\clearpage
\printindex

\end{document}
