\documentstyle[11pt,alltt,fullpage,html]{report}

\newcommand{\tcl}{{\sf Tcl}}
\newcommand{\tk}{{\sf Tk}}
\newcommand{\caml}{{\sf Caml}}
\newcommand{\camltk}{{\sf CamlTk}}
\newcommand{\wish}{{\sf wish}}
\newcommand{\thisrelease}{{\tt alpha3}}
\title{The \caml-\tk\ interface \\
       Release \thisrelease}
\author{Fran\c{c}ois Pessaux and Fran\c{c}ois Rouaix \\
           Projet Cristal, INRIA Rocquencourt \\
        {\small Francois.Pessaux@inria.fr, Francois.Rouaix@inria.fr}
        }
\date{November 1994}
\begin{document}
\maketitle


\section*{Introduction}
\tk\ is a GUI (graphical user-interface) library for the \tcl\ language.
Normally, \tk\ is controlled from \tcl. The purpose of the \camltk\
interface is to provide a \caml\ library to control \tk\ from \caml\
programs. Thus, \tk\ can be used to program user-interfaces in \caml\
without knowledge of the \tcl\ language.

\section*{Installation}

\begin{itemize}
\item This release runs only on Unix systems.
\item You need Caml Light 0.61.
\item You need to have \tcl\ and \tk\ installed. This release requires that
you have installed the includes and libraries, and \wish.
This release has been mostly tested with \tk\ 3.6.
\end{itemize} 

\begin{htmlonly}
First get the source
\htmladdnormallink{here}{ftp://ftp.inria.fr/lang/caml-light/camltk-alpha3.tar.Z},
and extract the archive in the {\tt contrib} directory of the Caml Light
distribution. 
Check 
\htmladdnormallink{here}
 {http://pauillac.inria.fr/\%7Erouaix/camltk-porting.html}
for problems reported since the distribution.
\end{htmlonly}

See the INSTALL file for instructions regarding the installation.


\section*{This documentation}
{\bf The present documentation is still sketchy, and some basic background
about toolkits, event-driven programming, and \tk\ is assumed.}
Documentation for \tk\ is currently available in \cite{ouster94},
\cite{welch94}, and in the {\tt man} pages of the \tk\ distribution.
The distribution also includes a first shot at an \caml\ browser, which may
help in finding out the types of functions available in the library as well as
the type definitions, all this in an {\em hypertext} tool.


What you will find here is:
\begin{description}
\item[The user's manual] : describes how to write \caml\ programs using the
\tk\ interface and to compile them. 
\item[The implementor's manual] : describes how the interface library is
generated from a high-level description. This part is of interest mainly for
expert \tk\ users who wish to extend the interface to support more
functions, more widgets, or simply to fix bugs in test versions.
\item[The Tentative reference manual] : is simply a the collection of
interface files from the library, restricted to functions (i.e. type
descriptions do not appear). 
\end{description}

\section*{Bug reports}
Before reporting a bug, please check the \verb|KnownBugs| file in the
distribution directory.
For the time being, please report problems and bugs directly to
\verb|Francois.Rouaix@inria.fr| instead of \caml\ mailing lists.
Don't hesitate to make comments and suggestions, or ask questions.
Contributions (description of missing stuff) are welcome.

%\section*{Thanks}

\tableofcontents
\include{user}
\include{impl}

% Not included in html version
\include{reference}

\begin{htmlonly}
\chapter{Reference manual}
{\Large WARNING: The on-line reference manual corresponds to the version
being developped NOW. It may not be accurate for the last released version}

\htmladdnormallink{This}{/bin/psearch/doc-camltk/library}
is the on-line version.
\end{htmlonly}



\bibliography{tk}
\bibliographystyle{plain}
\appendix
\include{bnf}
\end{document}

