\documentstyle[11pt,alltt,html,fullpage]{report}

% Various notations
\newcommand{\tcl}	{{\sf Tcl}}
\newcommand{\tk}	{{\sf Tk}}
\newcommand{\caml}	{{\sf Caml Light}}
\newcommand{\camltk}	{{\sf CamlTk}}
\newcommand{\wish}	{{\sf wish}}
\newcommand{\thisrelease}{{\tt beta1 (for 0.7)}}


\title{The \camltk\ interface \\
       Release \thisrelease}

\author{Fran\c{c}ois Pessaux and Fran\c{c}ois Rouaix \\
           Projet Cristal, INRIA Rocquencourt \\
        {\small Francois.Pessaux@inria.fr, Francois.Rouaix@inria.fr}
        }
\date{December 1994}

\begin{document}
\maketitle
\tableofcontents
\newpage

\section*{Introduction}
\addcontentsline{toc}{section}{Introduction}

\tk\ is a GUI (graphical user-interface) library for the \tcl\ language.
Normally, \tk\ is controlled from \tcl. The purpose of the \camltk\
interface is to provide a \caml\ library to control \tk\ from \caml\
programs. Thus, \tk\ can be used to program user-interfaces in \caml\
without knowledge of the \tcl\ language.

\section*{Installation}
\addcontentsline{toc}{section}{Installation}

\begin{itemize}
\item This release runs only on Unix systems.
\item You need Caml Light 0.7.
\item You need to have \tcl\ and \tk\ installed. This release requires that
you have installed the includes and libraries, and \wish.
This release has been tested with \tcl\ 7.3 and \tk\ 3.6, but {\bf not} with
beta versions of \tk\ 4.0.
\end{itemize} 

\begin{htmlonly}
Check 
\htmladdnormallink{here}
                  {http://pauillac.inria.fr/\%7Erouaix/camltk-porting.html}
for problems reported since the distribution.
\end{htmlonly}

\section*{This documentation}
\addcontentsline{toc}{section}{This documentation}

{\bf The present documentation is still sketchy, and some basic background
about toolkits, event-driven programming, and \tk\ is assumed.}
Documentation for \tk\ is currently available in \cite{ouster94},
\cite{welch94}, and in the {\tt man} pages of the \tk\ distribution.
What you will find here is:
\begin{description}
\item[The user's manual] : a short tutorial describing how to write \caml\
programs using the \tk\ interface and how to compile them. 
\item[The implementor's manual] : describing how the interface library is
generated from a high-level description. This part is of interest mainly for
expert \tk\ users who wish to extend the interface to support more
functions, more widgets, or simply to fix bugs in test versions.
\item[The Tentative reference manual] : the collection of interface files
from the library.
\end{description}

The \camltk\ distribution also includes a first shot at a \caml\ browser,
an {\em hypertext} tool that may help in finding out the types of functions
available in the library as well as the type definitions, 

\section*{Bug reports}
\addcontentsline{toc}{section}{Bug reports}

Before reporting a bug, please check the \verb|KnownBugs| file in the
\verb|doc| directory of the distribution.
Please report problems and bugs directly to \verb|Francois.Rouaix@inria.fr|
instead of \caml\ mailing lists.
Don't hesitate to make comments and suggestions, or ask questions.
Contributions (description of missing stuff) are welcome.

\section*{Acknowledgments}
Thanks to Loic Prylli \verb|<Loic.Prylli@lip.ens-lyon.fr>| \newline
      and Christophe Raffalli \verb|<raffalli@cs.chalmers.se>|
for bug reports.

\include{user}
\include{impl}

% Not included in html version, by command line option
\include{reference}

\begin{htmlonly}
\chapter{Reference manual}
{\Large WARNING: The on-line reference manual corresponds to the version
being developped NOW. It may not be accurate for the last released version}

\htmladdnormallink{This}{/bin/psearch/doc-camltk/library}
is the on-line version.
\end{htmlonly}

\bibliography{tk}
\bibliographystyle{plain}
\appendix
\include{bnf}
\end{document}

