\chapter{User's Manual}
{\bf This manual is a simple tutorial for \camltk }.  When writing your own
programs, you should consult the reference manual
(chapter~\ref{chap:ref}), as well as the \tk\ man pages.  Whenever the
reference manual is unsufficient, you can check the \verb|Widgets.src| file
(which syntax is given in chapter~\ref{chap:impl}) since this file the
mappings between \tk\ and \caml\ for functions and data. 

\section{Modules}
The \camltk\ interface is provided as a \caml\ library, composed of the
following modules:

\begin{itemize}
\item \verb|tk|, containing the initialisation functions, some frequently
used functions, and all types manipulated by the library.

\item a module for each widget class (e.g. \verb|button|). The creation
function and the commands specific to this class are contained in the
module. The module bears the same name as the widget class, except for
\verb|toplevel| widgets which are defined in module \verb|toplevelw|.

\item various other modules grouping functions of a same family
(e.g. \verb|focus|, \verb|winfo|). 

\item the \verb|textvariable| module, for the \verb|-textvariable| option
support. 

\item \verb|protocol| and \verb|support|, which contain the internals
of the interface. For example, the abstract type of widgets is defined in
\verb|support|, as well as some widget naming utilities.
\end{itemize} 

All modules required by the \camltk\ interface are in the \verb|tklib.zo|
or \verb|tklib2.zo| library archives (only one of them is required,
depending on the architecture you choose, cf. section~\ref{sec:compiling}).

The organization of the interface is designed so that, in general, you
should \verb|#open| only the \verb|tk| module, and for the 
rest of the functions use the \verb|modulename__symbol| syntax (e.g.
\verb|text__insert|).
This convention is preferred because some functions (such as \verb|create|
and \verb|configure|) exist in many different modules. In this context,
referring to functions with exact qualified names makes programs clearer and
less error-prone.


\section{Basics}
As usual with toolkits, the simplest way to learn how to write programs is
by copy-paste-ing from available source code. In general, a \caml\ program
using the \tk\ toolkit will have the following structure :

\input helloworld.tex

The first step is to initialize the interface with \verb|OpenTk|, and this
returns the {\em toplevel} widget\footnote{Note that attempting to call a
\camltk\ function without previously having initialised the interface will
probably cause your program to core dump.}.
Then, the program creates some widgets, as children of the {\em toplevel}
widget. These widgets form the user interface of your application. In this
simple case, we have only one button, whose behaviour is to print a simple
message on the standard output. Finally, the program enters the
\verb|MainLoop|, and user-interaction takes place.

\section{Example}
Here is a complete working example with a few widgets and callbacks.
This program may be found in the \verb|test| directory of the distribution,
under the name \verb|addition.ml|.

\input addition.tex

The \verb|ScrollCommand| callback has the property of being called each time
the insertion cursor blinks in an \verb|entry| widget. This property is
responsible for the ``electric'' behavior of this program. Another approach
would have been to associate an action to the event ``key return has been 
pressed''.

\section{Widgets}
The description of widget commands and options is given in the reference
manual (chapter~\ref{chap:ref}). The semantics of commands is however not
described there, so the reader should also refer to the \verb|man| pages of
the \tk\ distribution. 
As of this release, all widgets of \tk\ 3.6 are supported, although some
commands might not be implemented. Remember also that only few of them have
been tested.

\section{Bindings}
\tk\ offers, in addition to widget-specific callbacks, a general facility
for binding functions to events. (bind(n)).
Bindings involves the following types, defined in the \verb|tk| module:
\input bindtypes.tex

Binding are manipulated with the function
\begin{verbatim}
bind: Widget -> (Modifier list * XEvent) list -> BindAction -> unit
\end{verbatim} 

The first argument of type \verb|Widget| is naturally the widget for which
this binding is defined.
The second argument of type \verb|(Modifier list * XEvent) list| is the 
succession of events that will trigger the callback.
The last argument is the action : \verb|BindRemove| will remove a binding,
\verb|BindSet| will set a binding, and \verb|BindExtend| will extend a
binding.

In many cases, a callback needs information about the event that
triggered it, such as the precise key that was pressed, the position of the
mouse, etc. Therefore, a callback is of type \verb|EventInfo -> unit|, 
and the callback will be invoked with argument the record containing
event information. Since event information can be quite large,
the user has to specify, as an \verb|EventField list|, which fields in this
record are relevant for this particular callback. 
The other fields will {\bf not} contain accurate information. 
Check also the \tk\ documentation for finding out which fields are valid for a
given event.

\paragraph{Example}
\begin{verbatim}
bind myCanvas [[],Motion] 
   (BindSet([Ev_MouseX; Ev_MouseY;Ev_Place;Ev_SendEvent;Ev_KeySymString], 
            mouse_moved)) ;;
\end{verbatim} 
will trigger the callback \verb|mouse_moved| when the event \verb|Motion|
occurs in the \verb|myCanvas| widget. The function \verb|mouse_moved| may
assume that the  information in the fields {\tt Ev\_MouseX, Ev\_MouseY,
Ev\_Place, Ev\_SendEvent, Ev\_KeySymString} of its argument contain accurate
information.

\paragraph{Text and Canvas bindings}
\mbox{}\newline

Bindings on objets in \verb|text| and \verb|canvas| widgets obey the same
rules, and the functions are 
\begin{verbatim}
tk__text_tag_bind : Widget -> TextTag -> 
      (Modifier list * XEvent) list -> BindAction -> unit
tk__canvas_bind : Widget -> TagOrId -> 
      (Modifier list * XEvent) list -> BindAction -> unit
\end{verbatim} 
Additional arguments are respectively the {\em tag} and the {\em tag or id}
for which the binding is defined.

\paragraph{Class bindings}
\mbox{}\newline

Binding for classes of widgets are managed with
\begin{verbatim}
tk__class_bind : string -> 
      (Modifier list * XEvent) list -> BindAction -> unit
\end{verbatim} 
The first argument should be a widget class name (e.g. \verb|"Button"|).
Widget class names in \camltk\ are the exact same names as in \tk.

\section{Exiting}
The function
\begin{verbatim}
tk__CloseTk : unit -> unit
\end{verbatim} 
will terminate the main interaction loop (\verb|tk__MainLoop|) and
destroy all widgets.

% You should always try to call \verb|CloseTk| before exiting. Otherwise, in
% the {\em two-process} architecture, the slave {\tt wish} may stay around for
% a while, as well as named pipes created by the communication protocol in
% \verb|/tmp|. In particular, something like  
% \begin{verbatim}
% signal SIGINT (Signal_handle (function _ -> CloseTk(); exit 0))
% \end{verbatim} 
% should help your application behave nicely. Naturally your own cleanup
% should be inserted also.

\section{Errors}
The \camltk\ interface may raise the following exceptions:
\begin{description}
\item[TkError of string]: this exception is raised when a \tcl\tk\ command
fails during an evaluation. Normally, static typing in \caml, in addition to
some run-time verifications should prevent this situation from happening.
However, some errors are due to the external environment (e.g.
\verb|selection__get|) and cannot be prevented. The exception carries the
\tcl\tk\ error message.

\item[IllegalWidgetType of string]:
this exception is raised when a widget command is applied to a widget of a
different or unappropriate class (e.g. \verb|button__configure| applied to
a {\em scrollbar} widget). The exception carries the class of the faulty
widget.

\item[Invalid\_argument of string]:
this exception is raised  when some data cannot be exchanged between \caml\
and \tk. This situation occurs for example when an illegal option is used
for creating a widget. However, this error may also be caused by a faulty or
incomplete description of a widget, widget command or type.
\end{description} 

\section{More examples}
The \verb|test| directory of the distribution contains several test
examples. 
The \verb|books-examples| directory contains translations in \camltk\ of
several examples from \cite{ouster94}.

%The \verb|tktop| directory contains a more advanced example, although
%not very robust.

{\bf All examples included in this distribution  should not be considered as
working applications. They are provided for documentation purposes only}.

\section{The Caml Browser}
In the directory {\tt browser} you will find a yet more complete example,
featuring many different kinds of widgets. The browser mimicks the 
\caml\ toplevel behaviour: you can open and close modules ({\tt \#open} and
{\tt \#close} directives), or add a library path ({\tt \#directory} directive).
You can enter a symbol (value, type constructor, value constructor,
exception), and see its definition. You can also see the complete interface
of a module by double-clicking on it. 
Remember that the browser will show you a pretty-printed version
of compiled interfaces ({\tt .zi} files), so you will not see the comments
that were in the source file.

The browser can also display source files. Hypertext navigation is also
available for source files, but does not reflect the compiler's semantics,
as the \verb|#open| directives in the source file are not taken into
account.


\section{Compilation}
\label{sec:compiling}
\subsection{Architecture}
In this release, there are two possible architectures for the \camltk\
interface. As in previous versions, there is a {\em two-process}
architecture, where \caml\ communicates to a slave \wish\ interpreter.
There is now a {\em single-process} architecture, where both \caml\ and \tk\
coexist in a single process.

{\bf Note: } because the {\em two-process} architecture is less portable and
less robust than the {\em single-process} architecture, the former may not
be supported in future versions.


%% Avec la 0.7, Camltk est installe avec le reste du bouzin
% In the following, we refer to \verb|CAMLTKDIR| as an environment variable
% pointing to the directory where the interface library has been
% installed, that is normally \verb|/usr/local/lib/caml-light/tk|.
In the following, we refer to \verb|TCLLIBDIR| (resp. \verb|TKLIBDIR|) as
an environment variable containing the directory where \verb|libtcl.a|
(resp. \verb|libtk.a|) is located. Consult your system administrator if you
don't have this information. Moreover, we assume a ``standard installation''
of \caml\ and \camltk\ from the distribution. This means, starting from
release \thisrelease, that all \camltk\ library files are installed in the
same directory as standard \caml\ library files.

\subsection{Compilation for Single-process architecture}
The library is named \verb|tklib.zo|.
The usual commands for compiling and linking are:
\begin{verbatim}
$ camlc -c addition.ml
$ camlc -custom -o addition tklib.zo addition.zo \
        -ccopt -L$TCLLIBDIR -ccopt -L$TKLIBDIR \
        -lcamltk -lcaml -ltk -ltcl -lX11
\end{verbatim}
Linking is a bit complex, because we need to tell the C compiler/linker
where to find the libtk.a and libtcl.a libraries, and to link with all
required pieces. Your linker might complain about \verb|main| being defined
twice. It this produces an error, you cannot use this architecture.
We will try to provide a bug-fix if this turns out to be a problem.

\subsection{Compilation for Two-process architecture}
The library is named \verb|tklib2.zo|.
The usual commands for compiling and linking are:
\begin{verbatim}
$ camlc -c addition.ml
$ camlc -custom -o addition unix.zo tklib2.zo addition.zo -lunix
\end{verbatim} 
 \verb|unix.zo| comes from the \verb|libunix| contrib, and is normally
installed in the same directory as standard \caml\ libraries. 
The commands are simpler because we rely on the external \verb|wish|
interpreter.

\subsection{Toplevels}
The distribution also contains two toplevels featuring the \camltk\
interface: \verb|camltktop1| and \verb|camltktop2|.
These toplevels may be run with: 
\begin{verbatim}
$ camllight camltktop1
$ camllight camltktop2
\end{verbatim} 
Note however that the usage of the toplevels is awkward (for the time
being), because callbacks cannot be executed until you enter MainLoop, and
you cannot easily leave MainLoop.

\section{Extensions}
\camltk\ has already one extension (i.e. a feature not normally available in
\tk\ 3.6). This extension allows the association of callbacks to Unix input
file descriptors.

\subsection{File descriptor callbacks}
A callback can be associated to a file descriptor using the following
primitive:
\begin{verbatim}
tk__add_fileinput : file_descr -> (unit -> unit) -> unit
\end{verbatim} 
When some input is available on the file descriptor specified by the first
argument, then the callback (second argument) is called.
\begin{verbatim}
tk__remove_fileinput : file_descr -> unit
\end{verbatim} 
Removes the file descriptor callback. 

It is the programmer's responsability
to check for end of file, and to remove the callback if the file descriptor
is closed. Be aware that calling the \verb|update| function will potentially
destroy the sequentiality of read operations. Finally, note that during the
invocation of your callback, only one call to \verb|read| is guaranteed to
be non-blocking.

\section{Translating \tk\ idioms}
If you are a seasoned \tk\ programmer, you will find out that some \tk\
idioms have a different form in \camltk\ .

\subsection{Widgets}
First of all, widget creation is more ``functional''. One does not specify
the name of the created widget (except for \verb|toplevel| widgets), but
only its parent. The name is allocated by the library, and a handle to the
widget is returned to you. Then, widgets are not ``procedures''. They are
objects, and must be passed to widget manipulation functions.
For example,
\begin{verbatim}
button .myf.bok -title "Ok" -relief raised
\end{verbatim} 
is translated by (assuming that \verb|myf| is the parent widget \verb|.myf|)
\begin{verbatim}
let b = button__create myf [Title "Ok"; Relief Raised] in
...
\end{verbatim} 
Then, 
\begin{verbatim}
.myf.bok configure -state disabled
\end{verbatim} 
would be in \caml:
\begin{verbatim}
button__configure b [State Disabled]
\end{verbatim} 

This is more in the spirit of ML, but unfortunately eliminates the
possibility to specify configuration of widgets by resources based on widget 
names. It you absolutely want to use resources then you may call the
alternate \verb|create_named| function:
\begin{verbatim}
let b = button__create_named myf "bok" [Title "Ok"; Relief Raised] in
...
\end{verbatim} 
Assuming that \verb|myf| is the widget of path \verb|.myf|, then \verb|b|
will have path \verb|.myf.bok|. As in \tk, it is your responsibility to use
valid identifiers for path elements.

When widgets are mutually recursive (through callbacks, for example when
linking a scrollbar to a text widget), one should first create the widgets,
and then set up the callbacks by calling \verb|configure| on the widgets.
This is in contrast with \tcl\tk\ where one may refer to a procedure that
has not yet been defined.

Partially applied widget commands (such as redisplay commands) translate
quite well to \caml, with possibly some wrapping required due to 
value constructors.

\subsection{Text variables}
\verb|TextVariables| are now available in \camltk. However,
\verb|textvariables| are not physically shared between \caml\ and \tk.
Instead, setting or consulting a \verb|textvariable| requires a transaction
between \caml\ and \tk. However, the "sharing" properties of 
\verb|textvariables| in \tk\ are preserved. For example, if the
\verb|textvariable| is  
used to physically share the content of an {\tt entry} with the text
of a {\tt label}, then this sharing will effectively occur in the \tk\
world, whether your \caml\ program is concerned with the actual
contents of the \verb|textvariable| or not. 

\subsection{How to find out the \caml\ name of \tk\ keywords and functions}
Normally, a \caml\ command for a widget class will have the same name as the
corresponding \tk\ command. However, in some cases, for example when a same
command is used to read and set values, or has a variable number of type
incompatible arguments, \camltk\ will have several versions of the command
with different names.

For the time being the best information sources are:
\begin{itemize}
\item  the .mli files in the \verb|lib| directory (they also appear in
chapter~\ref{chap:ref}). This will give you at least the functions for each
widget class. 

\item  the \verb|Widgets.src| file of the distribution. 

\end{itemize} 

\section{Debugging}
Since this is a beta release, some \tk\ functions may have been improperly
implemented in the \camltk\ library. This may cause undue \tk\ errors
(exception \verb|TkError|, or sometimess \verb|Invalid_Argument| ). To
facilitate the debugging, you can set the 
Unix environment variable \verb|CAMLTKDEBUG| to any value before launching
your program. This will allow you to see all
data transferred between the \caml\ and the \tk\ processes. Since this data is
essentially \tcl\tk\ code, you need a basic knowledge of this language to
understand what is going on.
It is also possible to trigger debugging by setting the boolean reference
\verb|protocol__debug| to true (or false to remove debugging).
