\chapter{User's Manual}
{\bf This manual is a simple tutorial for \caml\tk}.  When writing your own
programs, you should consult the partial reference manual , as well as the
\tk\ man pages.  Whenever the reference manual is unsufficient, you can
check the \verb|Widgets.src| file (which syntax is given in
chapter~\ref{chap:impl}) since it describes the mappings between
\tk\ and \caml\ for functions and data. 

\section{Modules}
All modules required by the \tk\ interface are in the \verb|tklib.zo| 
library archive. These modules are
\begin{itemize}
\item \verb|tk|, containing the initialisation functions, some frequently
used functions, and all types manipulated by the library.

\item a module for each widget class (e.g. \verb|button|). The creation
function and the commands specific to this class are contained in the
module.

\item various other modules grouping functions of a same family
(\verb|focus|, \verb|winfo|). In general, you should not \verb|#open| these
files since many functions bear the same names in different modules.
Instead, you should use the \verb|modulename__symbol| syntax.

\item the \verb|textvariable| module, for the \verb|-textvariable| option
support.

\item \verb|protocol| and \verb|support|, which contain the internals
of the interface, and are of no interest to the programmer.
\end{itemize} 

\section{Basics}
As usual with toolkits, the simplest way to learn how to write programs is
by copy-paste-ing from available source code. In general, a \caml\ program
using the \tk\ toolkit will have the following structure :
\begin{verbatim}
(* Initialisation of the interface )
let top = tk__OpenTk ();;
(* top is now the toplevel widget *)
(* Widget initialisation *)
let b = button__create top ...
;;
(* Mainloop for events *)
tk__MainLoop()
;;
\end{verbatim} 

The first step is to initialize the interface with \verb|OpenTk|.
Then, the program creates some widgets, forming the user interface of your
application. Finally, the program enters the \verb|MainLoop|, and
user-interaction takes place.

\section{Example}
Here is a complete working example with a few widgets and callbacks.
This program may be found in the \verb|test| directory of the distribution,
under the name \verb|addition.ml|.

\begin{verbatim}
(* Make main functions and types available *)
#open "tk";;

let main () =
  let top = OpenTk ()  in
  (* The widgets. They all have "top" as parent widget.  *)
  let en1 = entry__create top [TextWidth 6; Relief Sunken] in
  let lab1 = label__create top [Text "plus"] in
  let en2 = entry__create top [TextWidth 6 ; Relief Sunken] in
  let lab2 = label__create top [Text "="] in
  let result_display = label__create top [] in
  (* References holding values of entry widgets *)
  let n1 = ref 0
  and n2 = ref 0  in
  (* Refresh result *)
  let refresh () =
    label__configure result_display [Text (string_of_int (!n1 + !n2))]  in
  (* Electric *)
  let get_and_refresh (w,r) =
    fun _ _ _ _ ->
      try
       r := int_of_string (entry__get w);
       refresh ()
      with
        Failure "int_of_string" ->
          label__configure result_display [Text "error"]
  in
  (* Set the callbacks *)
  entry__configure en1 [ScrollCommand (get_and_refresh (en1,n1)) ];
  entry__configure en2 [ScrollCommand (get_and_refresh (en2,n2)) ];
  (* Map the widgets *)
  pack [en1;lab1;en2;lab2;result_display] [];
  (* Make the window resizable *)
  wm__minsize_set top 1 1;
  (* Start interaction (event-driven program) *)
  MainLoop ()
;;

printexc__f main () ;;
\end{verbatim}


\section{Widgets}
The full description of widgets, their creation options, their commands, is
out of the scope of this document. The reader should refer to the \verb|man|
pages of the distribution. The commands for a given widget class are in 
a module bearing the same name as the class (i.e. \verb|button|), except
for {\em toplevel} widgets, which are defined is \verb|toplevelw| (due to a 
technical problem with the Caml Light toplevel).
As of this release, all widgets of \tk3.6 are supported, although some
commands might not be implemented. The most glaring omission is the {\tt
placer} geometry manager.

\section{Bindings}
\tk\ offers, in addition to widget-specific callbacks, a general facility
for binding functions to events. (bind(n)).
In the \caml\tk\ interface, this is implemented by the function
\begin{verbatim}
bind: Widget -> (Modifier list * XEvent) list -> BindAction -> unit
\end{verbatim} 
with the following types
\begin{verbatim}
type Modifier =
    Control
  | Shift
  | Lock
  | Button1
  | Button2
  | Button3
  | Button4
  | Button5
  | Any
  | Double
  | Triple
  | Mod1
  | Mod2
  | Mod3
  | Mod4
  | Mod5
  | Meta
  | Alt 
;;

type XEvent =
    XKey of string	(* /usr/include/X11/keysymdef.h *)
  | ButtonPress
  | Button
  | ButtonRelease
  | Circulate
  | CirculateRequest
  | ColorMap
  | Configure
  | ConfigureRequest
  | Destroy
  | Enter
  | Expose
  | FocusIn
  | FocusOut
  | Gravity
  | Keymap
  | KeyPress
  | Key
  | KeyRelease
  | MapRequest
  | Motion
  | Leave
  | Map
  | Property
  | Reparent
  | ResizeRequest
  | Unmap
  | Visibility 
  | WhatButton of int
;;

type EventField =
    Above
  | ButtonNumber
  | Count
  | Detail
  | Focus
  | Height
  | KeyCode
  | Mode
  | State
  | ValueMask
  | Width
  | MouseX
  | MouseY
  | WidgetName
  | Time 
  | OverrideRedirect
  | Place
  | Char
  | BorderWidth
  | Display
  | SendEvent
  | KeySymString
  | KeySymInt
  | RootWindow
  | SubWindow
  | Type
  | RootX
  | RootY
;;

type EventInfo = {
  mutable Ev_Above : int;
  mutable Ev_ButtonNumber : int;
  mutable Ev_Count : int;
  mutable Ev_Detail : string;
  mutable Ev_Focus : bool;
  mutable Ev_Height : int;
  mutable Ev_KeyCode : int;
  mutable Ev_Mode : string;
  mutable Ev_State : string;
  mutable Ev_ValueMask : int;
  mutable Ev_Width : int;
  mutable Ev_MouseX : int;
  mutable Ev_MouseY : int;
  mutable Ev_WidgetName : string;
  mutable Ev_Time : string;
  mutable Ev_OverrideRedirect : string;
  mutable Ev_Place : string;
  mutable Ev_Char : string;
  mutable Ev_BorderWidth : int;
  mutable Ev_Display : int;
  mutable Ev_SendEvent : int;
  mutable Ev_KeySymString : string;
  mutable Ev_KeySymInt : int;
  mutable Ev_RootWindow : int;
  mutable Ev_SubWindow : int;
  mutable Ev_Type : int;
  mutable Ev_RootX : int;
  mutable Ev_RootY : int }
;;

type BindAction =
   BindSet of EventField list *  (EventInfo -> unit)
 | BindRemove
;;
\end{verbatim} 
The first argument of type \verb|Widget| is naturally the widget for which
this binding is defined (there is no support yet for bindings on whole
widget classes).
The second argument of type \verb|(Modifier list * XEvent) list| is the 
succession of events that will trigger the callback.
The last argument is the action : \verb|BindRemove| will remove a binding.
\verb|BindSet| will set a binding.
The argument of \verb|BindSet| are an \verb|EventField list| and the
callback itself. 
In bindings, the callback needs some information about the event that
triggered it. This information is contained in the \verb|EventInfo| passed
to the callback, but the user has to specify which fields in this record are
relevant for this particular callback. The other fields do {\bf not} contain
accurate information. Check also the \tk\ documentation for finding out
which field is valid for a given event.

\paragraph{Example}
\begin{verbatim}
bind MyCan [[],Motion] 
   (BindSet([MouseX; MouseY;Place;SendEvent;KeySymString], BindCmd)) ;;
\end{verbatim} 
will trigger the callback \verb|BindCmd| when the event \verb|Motion| occurs
in the \verb|MyCan| widget. The command \verb|BindCmd| may assume that the 
information in the fields {\tt Ev\_MouseX, Ev\_MouseY, Ev\_Place, Ev\_Place,
Ev\_SendEvent, Ev\_KeySymString} of its \verb|EventInfo| argument contain
accurate information.

Bindings for \verb|text| and \verb|canvas| widgets obey the same rules,
and the functions are 
\begin{verbatim}
tk__text_tag_bind : Widget -> TextTag -> 
      (Modifier list * XEvent) list -> BindAction -> unit
tk__canvas_bind : Widget -> TagOrId -> 
      (Modifier list * XEvent) list -> BindAction -> unit
\end{verbatim} 
Additional arguments are respectively the {\tt tag} and the {\tt tag or id}
for which the binding is defined.

\section{Exiting properly}
You should always try to call \verb|CloseTk| before exiting. Otherwise, the
slave {\tt wish} may stay around for a while, as well as named pipes created
by the communication protocol in \verb|/tmp|. In particular, something like 
\begin{verbatim}
signal SIGINT (Signal_handle (function _ -> CloseTk(); exit 0))
\end{verbatim} 
should help your application behave nicely. Naturally your own cleanup
should be inserted also.

\section{More examples}
The \verb|test| directory of the distribution contains several other
examples. Read them to learn more about \caml\tk\ programming.
The \verb|tktop| directory contains a more advanced example, although
nor very robust.

{\bf All examples included in this distribution  should not be considered as
working applications. They are provided for documentation purposes only}.

\section{The Caml Browser}
In the directory {\tt browser} you fill find a yet more complete example,
featuring many different kinds of widgets. The browser mimicks the Caml
Light toplevel behaviour: you can add open and close modules ({\tt \#open} and
{\tt \#close} directives), or add a library path ({\tt \#directory} directive).
You can enter a symbol (value, type constructor, value constructor,
exception), and see its definition. You can also see the complete interface
of a module by double-clicking on it.
The browser works on compiled interfaces ({\tt .zi} files) only, not on
source files.

\section{Compilation}
The usual commands for compiling programs using the \caml\tk interface are
\begin{verbatim}
$ camlc -c -I $CAMLTKDIR addition.ml
$ camlc -custom -I $CAMLTKDIR -o addition unix.zo tklib.zo addition.zo -lunix
\end{verbatim} 
The \verb|CAMLTKDIR| environment variable should point to the directory where
the interface library has been installed, that is normally
\verb|/usr/local/lib/caml-light/tk|. \verb|unix.zo| comes from the
\verb|libunix| contrib, and is normally installed in the same directory as
standard \caml\ libraries. 

The distribution \verb|Makefile| also contains an entry to produce a \caml\
toplevel featuring the \tk\ interface : \verb|camltktop|.
This toplevel may be run with: 
\begin{verbatim}
$ camllight -I $CAMLTKDIR $CAMLTKDIR/camltktop
\end{verbatim} 

\section{Extensions}
The alpha versions have already one extension (i.e. a feature not
normally available in \tk\ 3.6). This extension allows the association of
callbacks to Unix input file descriptors. 

\subsection{File descriptor callbacks}
A callback can be associated to a file descriptor using the following
primitive:
\begin{verbatim}
tk__add_fileinput : file_descr -> (unit -> unit) -> unit
\end{verbatim} 
When some input is available on the file descriptor specified by the first
argument, then the callback (second argument) is called.
\begin{verbatim}
tk__remove_fileinput : file_descr -> unit
\end{verbatim} 
Removes the file descriptor callback. It is the programmer's responsability
to check for end of file, and to remove the callback if the file descriptor
is closed.

\section{Translating \tk\ idioms}
If you are a seasoned \tk\ programmer, you will find out that some \tk\
idioms have a different form in \camltk\ .

\subsection{Widgets}
First of all, widget creation is more ``functional''. One does not specify
the name of the created widget (except for \verb|toplevel| widgets), but
only its parent. The name is allocated by the library, and a handle to the
widget is returned to you. Then, widgets are not ``procedures''. They are
objects, and must be passed to widget manipulation functions.
For example,
\begin{verbatim}
button .myf.bok -title "Ok" -relief raised
\end{verbatim} 
is translated by
\begin{verbatim}
let b = button__create myf [Title "Ok"; Relief Raised] in
...
\end{verbatim} 
assuming that \verb|myf| is the parent widget \verb|.myf|.
Then, 
\begin{verbatim}
.myf.bok configure -state disabled
\end{verbatim} 
would be in \caml:
\begin{verbatim}
button__configure b [State Disabled]
\end{verbatim} 

When widgets are mutually recursive (through callbacks, for example when
linking a scrollbar to a text widget), one should first create the widgets,
and then set up the callbacks by calling \verb|configure| on the widgets.
Otherwise, partially applied widget commands (such as redisplay commands)
translate quite well to \caml, with possibly some wrapping required due to
value constructors.

\subsection{Text variables}
\verb|TextVariables| are now available in \caml\tk. However,
\verb|textvariables| are not physically shared between \caml\ and \tk.
Instead, setting or consulting a \verb|textvariable| requires a transaction
between \caml\ and \tk. However, if the \verb|textvariable| is used to
physically share the content of an {\tt entry} with the text of a {\tt
label}, then this sharing will effectively occur in the \tk\ world, and
\caml\ is not concerned with the actual contents of the \verb|textvariable|.

\subsection{How to find out the \caml\ name of \tk\ keywords and functions}
Since this documentation is not a reference manual, you have unfortunately
to make some efforts to find out what functions are available, and how the
datatypes are defined. For the time being the best methods are:
\begin{itemize}
\item Check the .mli files in the library directory (they also appear in
chapter~\ref{chap:ref}). This will give you at
least the functions for each widget class.
\item The types are almost all defined in the \verb|tk| module. Instead of
looking at the source, you can use the {\tt mletags} utility in conjunction
with the {\tt tags} mechanism of Emacs. 
\item Use the browser
\end{itemize} 

\section{Debugging}
Since this is an alpha release, some \tk\ functions may have been improperly
implemented in the \caml\ library. This may cause undue \tk\ errors
(exception \verb|TkError|). To facilitate the debugging, you can set the
Unix environment variable \verb|CAMLTKDEBUG| to any value before launching
your program. This will allow you to see all
data transferred between the \caml\ and the \tk\ processes. Since this data is
essentially \tcl\tk\ code, you need a basic knowledge of this language to
understand what is going on.
It is also possible to trigger debugging by setting the boolean reference
\verb|protocol__debug| to true (or false to remove debugging).


